Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics

### 2019-08-14 scagnostics.Rmd
adjustments to introduction

The term "projection pursuit" (PP) was coined by @FT74 to describe a procedure for searching high (say $p-$)dimensional data for "interesting" low-dimensional projections ($d=1$ or $2$ usually). The procedure, originally suggested by @kr69, involves defining a criterion function, or index, that measures the "interestingness" of each $d$-dimensional projection of $p$-dimensional data. This criterion function is optimized over the space of all $d$-dimensional projections of $p$-space, searching for both global and local maxima. It is hoped that the resulting solutions reveal low-dimensional structure in the data not found by methods such as principal component analysis. Projection pursuit is primarily used for visualization, with projected data always reported as plots.

A large number of projection pursuit indexes have been developed, primarily based on departure from normality, which includes clusters, outliers and skewness, and also for finding separations between known groups (e.g. @f87, @hall89 @CBC92, @naito1997, @lckl2005, @AHC02, @CEM:CEM2568, @JS87, @5508437, @PAN2000153, @FER13, @LOPERFIDO201842).  Less work has been done on indexes to find nonlinear dependence between variables, when $d>1$.

@@

The motivation is based on physics applications, to aid the interpretation of model fits on experimental results. A physical model can be considered to be a set of $p$ free parameters, that cannot be measured directly and are determined by fitting a set of $q (p<q)$ experimental observations, for which predictions can be made once the $p$ parameters are estimated. (Note here, that while we may have analytic expressions for the predictions, this is not always the case and we often have to rely on numerical computation.) Different sets of model parameters ($n$) found to be  compatible with the experimental results within a selected level of confidence yield the data to be examined using projection pursuit. A single prediction can be a complicated function of all of the free parameters, and typically $q \in [100,1000]$ and $p \sim 10$. Current practice is to examine pairs of parameters, or combinations produced by intuition or prior knowledge. This begs the question, whether important nonlinear associations are missed because tools to search are not available.

\textcolor{red}{PP can also be used to explore data with many more variables, data that is commonly seen today. For example, [@Cook:2018mvr] explores a 56-dimensional parameter space, by first reducing the number of dimensions to the first six principal components, before applying projection pursuit. There is not a single solution for working with high-dimensional data. PCA was appropriate for this problem because reducing to principal component space removed the linear dependencies but preserved the nonlinear relationships that were interesting to discover. Some projection pursuit indexes incorporation penalty terms to naturally deal with noise dimensions. Having an efficient optimizer can also important with increasing dimensionality, because there are many more dimensions to search.}

To find appropriate projections pursuit indexes, literature on variable selection was examined. With high-dimensional data, even plotting all pairs of variables can lead to too many plots, which is what "scagnostics" (@scag, @WW08) were developed to address by providing metrics from which to select the most interesting variable pairs. There are eight scagnostics, of which three ("convex", "skinny" and "stringy") are used here. The question is whether these can be adapted into projection pursuit indexes, to search for unusual features in two-dimensional projections of high-dimensional data. Recent PhD research by @Grimm2016 explored the behavior of scagnostics for selecting variables, and proposed two more that have nicer properties, based on smoothing splines and distance correlation. In addition, two more indexes for measuring dependence have been proposed in the machine learning literature, based on information criteria, maximal and total information coefficient (MIC and TIC) [@Reshef1518], with computationally more efficient versions (MIC_e, TIC_e) [@JMLRv1715308]. These are related to original 1D projection pursuit indexes based on entropy (e.g. @huber85, @JS87). This provides seven current indexes for measuring dependence between two variables, and each is available in an R [@rref] package: scagnostics [@HWscagR], mbgraphic [@mbgraphic] and minerva [@minerva].

Projection pursuit can be understood and investigated more substantially when combined with a tour. A tour (@As85, @BCAH05) displays a smooth sequence of low dimensional projections from high dimensions to explore multivariate data for structure such as clusters, outliers, and nonlinear relationships. @CBCH94 provided an approach combining the tour algorithm with projection pursuit, to interactively both search for interesting projections, and examine the behavior of the indexes. The projection pursuit guided tour is available in the R package, tourr [@tourr], and provides optimization routines, and visualization.

\textcolor{red}{This paper is structured as follows.} Section \ref{sec:construct} discusses index construction, and how they can be used in the guided tour. Section \ref{sec:investigate} investigates the behavior of the indexes, explored primarily using tour methods. The new guided tour with these indexes is applied to two examples from gravitational wave astronomy (Section \ref{sec:phys}). \textcolor{red}{The latter two parts are connected in that the application of the new indexes to these problems is the main motivation for the paper, and the simulation study, in the first part, was conducted to better understand the behavior of the indexes in general. The techniques in Section \ref{sec:investigate} define procedures that will be generally useful for researchers developing new projection pursuit indexes to assess their behavior. The paper finishes with a discussion about the limitations of this work, and the potential future directions.}

>>>>

The term "projection pursuit" (PP) was coined by @FT74 to describe a procedure for searching high (say $p-$)dimensional data for "interesting" low-dimensional projections ($d=1$ or $2$ usually). The procedure, originally suggested by @kr69, involves defining a criterion function, or index, that measures the "interestingness" of each $d$-dimensional projection of $p$-dimensional data. This criterion function is optimized over the space of all $d$-dimensional projections of $p$-space, searching for both global and local maxima. It is hoped that the resulting solutions reveal low-dimensional structure in the data not found by methods such as principal component analysis. Projection pursuit is primarily used for visualization, with the projected data always reported as plots.

A large number of projection pursuit indexes have been developed, primarily based on departure from normality, which includes clusters, outliers and skewness, and also for finding separations between known groups (e.g. @f87, @hall89 @CBC92, @naito1997, @lckl2005, @AHC02, @CEM:CEM2568, @JS87, @5508437, @PAN2000153, @FER13, @LOPERFIDO201842).  Less work has been done on indexes to find nonlinear dependence between variables, focused on $d=2$, which motivates this research.

@@

The driving application is from physics, to aid the interpretation of model fits on experimental results. A physical model can be considered to be a set of $p$ free parameters, that cannot be measured directly and are determined by fitting a set of $q~ (p<q)$ experimental observations, for which predictions can be made once the $p$ parameters are estimated. (Note here, that while we may have analytic expressions for the predictions, this is not always the case and we often have to rely on numerical computation.) Different sets of model parameters ($n$) found to be  compatible with the experimental results within a selected level of confidence yield the data to be examined using projection pursuit. A single prediction can be a complicated function of all of the free parameters, and typically $q \in [100,1000]$ and $p \sim 10$. Current practice is to examine pairs of parameters, or combinations produced by intuition or prior knowledge. This begs the question, whether important nonlinear associations are missed because because they are hidden in linear combinations of more than two variables.

\textcolor{red}{PP can be combined with other dimension reduction methods when $p$ is very high. For example, it can be beneficial to first do principal component analysis prior to PP, especially to remove linear dependencies before searching for other types of association. This is the approach used in @Cook:2018mvr, which explores a 56-dimensional parameter space, by first reducing the number of dimensions to the first six principal components, before applying projection pursuit. PCA was appropriate for this problem because reducing to principal component space removed the linear dependencies while preserving the nonlinear relationships that were interesting to discover. Some projection pursuit indexes do incorporate penalty terms to automate removing noise dimensions. It can also be important to have an efficient PP optimizer, particularly when working with high dimensions, because the search space increases exponentially with dimension.}

To find appropriate projections pursuit indexes for detecting nonlinear dependencies, the literature on variable selection was a starting point. With high-dimensional data, even plotting all pairs of variables can lead to too many plots, which is what "scagnostics" (@scag, @WW08) were developed to address by providing metrics from which to select the most interesting variable pairs. There are eight scagnostics, of which three ("convex", "skinny" and "stringy") are used here. The question is whether these can be adapted into projection pursuit indexes, to search for unusual features in two-dimensional projections of high-dimensional data. Recent PhD research by @Grimm2016 explored the behavior of scagnostics for selecting variables, and proposed two more that have nicer properties, based on smoothing splines and distance correlation. In addition, two more indexes for measuring dependence have been proposed in the machine learning literature, based on information criteria, maximal and total information coefficient (MIC and TIC) [@Reshef1518], with computationally more efficient versions (MIC_e, TIC_e) [@JMLRv1715308]. These are related to original 1D projection pursuit indexes based on entropy (e.g. @huber85, @JS87). This provides seven current indexes for measuring dependence between two variables, and each is available in an R [@rref] package: scagnostics [@HWscagR], mbgraphic [@mbgraphic] and minerva [@minerva].

PP index behavior can be understood and investigated more substantially when combined with a tour. A tour (@As85, @BCAH05) displays a smooth sequence of low dimensional projections from high dimensions to explore multivariate data for structure such as clusters, outliers, and nonlinear relationships. @CBCH94 provided an approach combining the tour algorithm with PP, to interactively both search for interesting projections, and examine the behavior of the indexes. The projection pursuit guided tour is available in the R package, tourr [@tourr], and provides optimization routines, and visualization.

\textcolor{red}{This paper is structured as follows.} Section \ref{sec:construct} discusses index construction, and how they can be used in the guided tour. Section \ref{sec:investigate} investigates the behavior of the indexes, explored primarily using tour methods. The new guided tour with these indexes is applied to two examples from gravitational wave astronomy (Section \ref{sec:phys}). \textcolor{red}{The latter two parts are connected in that the application of the new indexes to these problems is the main motivation for the paper, and the simulation study, in the first part, was conducted to better understand the behavior of the indexes in general. The techniques in Section \ref{sec:investigate} define procedures that will be generally useful for researchers developing new projection pursuit indexes to visually assess their behavior. Visual methods to diagnose the index behavior is important because PP is primarily used for visualistion.  The paper finishes with a discussion about the limitations of this work, and the potential future directions.}
